### 2020-10-28 latex/acrobot.tex
Added comments on what the proof should look like for the acrobot
First completion draft of acrobot constraint chapter

While this is the most general representation of an acrobot, the dynamics
become unwieldy.
To make rigorous analysis of these dynamics more tractable, we begin by assuming
the acrobot is comprised of two massless rods of equal length \(l\), with equal
point masses \(m\) at the tips.
We call this a \textit{simple} acrobot, which is displayed in Figure
\ref{fig:simple-acrobot-model}.
We will also ignore any frictional forces at both the hip and shoulder joints. 
Finally, it is important to note that a real gymnast cannot swing their legs in
full circles; 
for this reason, we assume that \(q_a \in [-Q_a, Q_a]\) where 
\(Q_a \in ]-\pi, \pi[\).

@@

\section{The Acrobot Constraint}
% TODO: Give the constraint, state the theorem, and show the outline of the
% proof (both as a sketch and in a figure)
One may be tempted to design a constraint \(q_a = f(\theta)\) 
with \(\theta := \arctan_2(p_u,q_u)\), since it was so effective for the VLP in
Chapter \ref{ch:vlp}.
It can be challenging to show this type of constraint is regular.
Recall that all constraints of the form \(h(q,p) = q_a - f(p_u)\)
are regular, since the regularity equation (\ref{eqn:regularity-matrix-acrobot})
becomes strictly positive.
A momentum-based constraint approximates human motion may still be likely to
inject energy into the acrobot.

To design the acrobot constraint, let us imagine what a human does on
a seated swing.
When they are moving forward they extend their legs, and when they are moving
backward they retract them. 
In reality, the motion is slightly more complicated than that.:
as the swing gains speed, the person leans their body back while
extending their legs.
This allows them to bring their legs higher, shortening the distance
from their center of mass to the pivot and adding more energy to the swing.
When the swing moves backward, they sit up and fully retract their legs
underneath them.

Now imagine we have affixed the torso of the acrobot to the swing's rope, so 
that the legs can pivot both forwards and backwards.  
In this case, the legs would pivot in the direction of motion
proportionally to the swing's speed.
One VNHC which emulates this process is \(q_a = \bar{q}_a\arctan_2( I p_u)\), as
in Figure \ref{fig:qa-arctan}.
Here, \(\bar{q}_a \in ]0,\frac{2 Q_a}{\pi}]\) so that 
\(q_a \in [-Q_a, Q_a]\) and \(I > 0 \in \R\) is a fixed control parameter.

This constraint is not a perfect emulation of a gymnast's movement -- during
a giant the gymnast's legs are almost completely extended, while the constraint
retracts the legs partially during rotations.
However, the motion looks similar enough that this constraint should provide a
decent foundation for injecting energy into the acrobot.
It is for this reason that we choose our acrobot's constraint to be
\begin{equation}\label{eqn:acrobot-constraint}
    h(q,p) = q_a - \bar{q}_a \arctan_2(I p_u)
    .
\end{equation}

@@

Let us now compute the constrained dynamics for this
(\ref{eqn:acrobot-constraint}).
% TODO: should I do this for a generic h(q_a,p_u)? That's how I did it in my
% notes...
Note that \(dh_q = \begin{bmatrix}0 & 1\end{bmatrix}\tpose\), while
\[
    dh_{p_u} = \frac{-\bar{q}_a I}{1 + I^2 p_u^2}
    .
\]
Applying these to (\ref{eqn:g-qupu}), we get the solution for \(p_a\) on the
constraint manifold as
\[
    p_a(q_u,p_u) = \frac{
        (1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
    }{ml^2(3+2c_a)(1+I^2 p_u^2)}
    .
\]

\section{Proving the Acrobot Gains Energy}
% TODO: Prove this in sections

>>>>

While this is the most general representation of an acrobot, the dynamics
become unwieldy.
To make rigorous analysis of these dynamics more tractable, we begin by assuming
the acrobot is comprised of two massless rods of equal length \(l\), with equal
point masses \(m\) at the tips.
We call this a \textit{simple} acrobot, which is displayed in Figure
\ref{fig:simple-acrobot-model}.
We will also ignore any frictional forces at both the hip and shoulder joints. 
Finally, it is important to note that a real gymnast cannot swing their legs in
full circles, though they are usually flexible enough to raise them parallel to
the floor; 
for this reason, we assume that \(q_a \in [-Q_a, Q_a]\) where 
\(Q_a \in [\frac{\pi}{2}, \pi[\).

@@

\section{The Acrobot Constraint}
% TODO: Give the constraint, state the theorem, and show the outline of the
% proof (both as a sketch and in a figure)
One may be tempted to design a constraint of the form \(q_a = \bar{q}_a\sin(\theta)\) 
with \(\theta := \arctan_2(p_u,q_u)\), since a similar approach 
was so effective for the VLP in Chapter \ref{ch:vlp}.
Unfortunately, this constraint is not regular, and it is difficult to find any
VNHC of the form \(q_a = f(\theta)\) where it is easy to prove regularity
analytically.
Instead, we will develop a constraint \(h(q,p) = q_a - f(p_u)\) which,
as we saw earlier, is always regular and may still be likely to inject energy.

To design this constraint, let us imagine what a human does on
a seated swing.
When they are moving forward they extend their legs, and when they are moving
backward they retract them. 
In reality, the motion is slightly more complicated than that:
as the swing gains speed, the person leans their body back while
extending their legs.
This allows them to bring their legs higher, shortening the distance
from their center of mass to the pivot and adding more energy to the swing.
When the swing moves backward, they sit up and fully retract their legs
underneath them.

Now imagine we have affixed the torso of the acrobot to the swing's rope, so 
that the legs can pivot both forwards and backwards.
The legs cannot bend, so in an attempt to emulate motion the legs should pivot
in the direction of motion proportionally to the swing's speed.
One VNHC which emulates this process is \(q_a = \bar{q}_a\arctan_2( I p_u)\), as
in Figure \ref{fig:qa-arctan}.
Here, \(\bar{q}_a \in ]0,\frac{2 Q_a}{\pi}]\) so that 
\(q_a \in [-Q_a, Q_a]\) and \(I \in \R_{>0}\) is a fixed control parameter.

This constraint is not a perfect emulation of a gymnast's movement -- during
a giant the gymnast's legs are almost completely extended, while our constraint
retracts the legs partially during rotations.
However, the motion looks similar enough that this constraint should provide a
decent foundation for injecting energy into the acrobot.
It is for this reason that we choose our acrobot's constraint to be
\begin{equation}\label{eqn:acrobot-constraint}
    h(q,p) = q_a - \bar{q}_a \arctan_2(I p_u)
    .
\end{equation}

@@

Let us now compute the constrained dynamics under
(\ref{eqn:acrobot-constraint}).
% TODO: should I do this for a generic h(q_a,p_u)? That's how I did it in my
% notes...
Note that \(dh_q = \begin{bmatrix}0 & 1\end{bmatrix}\tpose\), while
\[
    dh_{p_u} = \frac{-\bar{q}_a I}{1 + I^2 p_u^2}
    .
\]
Applying these to (\ref{eqn:g-qupu}), we get the solution for \(p_a\) on the
constraint manifold as
\[
    p_a(q_u,p_u) = \frac{
        (1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
    }{ml^2(3+2c_a)(1+I^2 p_u^2)}
    .
\]
The dynamics for \(p_u\) do not contain \(p_a\), so they remain unchanged.
The constrained dynamics for \(q_u\) are given by 
\begin{equation*}
    \dot{q}_u = e_1\tpose \Minv(q) \begin{bmatrix}
                    p_u \\ p_a(q_u,p_u)
                \end{bmatrix} %\\
\end{equation*}
%        &= \frac{1}{ml^2(2-c_a^2)} 
%            \left( p_u - 
%                (1+c_a)\frac{(1+c_a)(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I (2-c_a^2)(2s_u + s_{ua})
%                }{ml^2(3+2c_a)(1+I^2 p_u^2)}
%            \right) 
%        ,
%\end{align*}
which one can simplify into 
\begin{equation*}
    \dot{q_u} = \frac{(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }{ml^2(1+I^2 p_u^2)(3+2c_a)}
    .
\end{equation*}
Hence, the constrained dynamics for the acrobot under
(\ref{eqn:acrobot-constraint}) are
\begin{equation}\label{eqn:acrobot-constrained-dynamics}
    \begin{cases}
        \dot{q}_u = \frac{(1+I^2 p_u^2)p_u - m^2gl^3\bar{q}_a I(2s_u + s_{ua})(1+c_a) }
            {ml^2(1+I^2 p_u^2)(3+2c_a)}
        , \\
        \dot{p}_u = - m g l (2s_u + s_{ua})
        .
    \end{cases}
\end{equation}

These dynamics do not always gain energy -- we need some additional conditions
on the control parameters to guarantee this is true. 
With these conditions, (\ref{eqn:acrobot-constraint}) injects energy into the
acrobot, as is explained by the following Theorem.

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot (\ref{eqn:acrobot-hamiltonian}) constrained by
    the regular VNHC \(h(q,p_u) = q_a - \bar{q}_a \arctan(I p_u)\).
    Suppose \(\bar{q}_a \in ]0,1]\).
    Then there exists a small enough\footnotemark \(I \in \R_{>0}\)
    such that the VNHC injects energy. 
    Setting \(I \in \R_{<0}\) allows the VNHC to dissipate energy.
\end{thm}
\footnotetext{In fact, we can be more precise than this. 
    Letting \(\epsilon > 0\), we need 
    \(I \leq \min\left\{I_1(\epsilon),I_2\right\}\) where
    \(I_1(\epsilon) = \frac{\tan(\epsilon)}{(\pi-\epsilon)}
    \left(\frac{m^2gl^3(\cos(\epsilon)-2\sin(\epsilon))}{\pi-\epsilon}
    \left(\frac{2\bar{q}_a\tan(\epsilon)}{(1+\tan(\epsilon)^2)(\pi-\epsilon)}+5\right) 
    \right)^{-1/2}\)
    and \(I_2 = \sqrt{\frac{3}{10 m^2gl^3\bar{q}_a^2}}\).
    See Apendix \ref{apx:I-bounds} for details.
}

Proving Theorem \ref{thm:acrobot-energy-stabilization} requires a change of
coordinates and the use of perturbation theory.
The full proof is given in Chapter \ref{sec:acrobot-proof}. 
We will conclude this section with a high level sketch of the proof.

The energy-based proof used in Chapter \ref{ch:vlp} does not work for the
acrobot, since no ``average energy" function has positive derivative everywhere
for these dynamics.
We instead use a perturbation analysis. 
When \(I = 0\), the acrobot behaves like a simple pendulum with two masses. 
There is a suitable change of coordinates into a pseudo-radius 
\(\mu \in \R_{>0}\) and pseudo-angle \(\alpha \in \mathbb{S}^1\) where \(\mu\)
is constant along orbits of the pendulum and \(\alpha\) is strictly increasing.
We can show there exists \(I > 0\) small enough such that
the orbit of the acrobot in \((\mu,\alpha)\) coordinates
tracks the pendulum motion to order \(I\).
What's more, we can show that \(\mu\) is increasing on average, which means the
acrobot is gaining energy.

\section{Proving the Acrobot Gains Energy}\label{sec:acrobot-proof}
% TODO: Prove this in sections
% TODO: You don't have to use the proof of theta_dot being positive. Stick to
% Manfredi's proof, and when you say "for I small enough" put in a footnote that
% says "in fact, we can be more precise and set bounds for I given as I <
% min{...}. See appendix for more detail."
% Then, we can have an appendix where we derive the bounds on I so that
% theta_dot is always positive. This will help the reader keep track and not
% bother them too much.

### 2020-10-28 latex/vlp.tex
First completion draft of acrobot constraint chapter

\begin{defn}\label{defn:energy-injection}
   Let \(D \subset \R^2\) be open with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

The second property of Definition \ref{defn:energy-injection} allows the system
to have unstable equilibria, but not limit cycles nor closed orbits.
The next definition ties this notion of energy gain to VNHCs.

\begin{defn}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects (dissipates) energy} if the constrained dynamics gain (lose)
   energy everywhere on the constraint manifold, except possibly on a set of
   measure zero.
\end{defn}

>>>>

\begin{defn}\label{defn:energy-gain}
   Let \(D \subset \R^2\) be open with \(0 \in D\).
   Let \(f : D \rightarrow \R^2\) be locally Lipschitz with \(f(0) = 0\).
   The system described by \(\dot{x} = f(x)\) 
   \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
   The system \textit{loses energy on \(D\)} if it gains energy in
   negative-time.
\end{defn}

The second property of Definition \ref{defn:energy-gain} allows the system
to have unstable equilibria, but not limit cycles nor closed orbits.
The next definition ties this notion of energy gain to VNHCs.

\begin{defn}\label{defn:energy-injection}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects (dissipates) energy} if the constrained dynamics gain (lose)
   energy everywhere on the constraint manifold, except possibly on a set of
   measure zero.
\end{defn}
